The effects of drought on land cover change and vegetation productivity in continental Chile

### 2024-05-09 article/paper_drought_Chile2023.qmd
update version may 2024

          country: 7500994, Chile

@@

          city:  Enschede, The Netherlands

@@

          country: Chile

@@

          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile, Santiago, Chile.

@@

@@

          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750, Santiago, Chile

@@

          name: Data Observatory Foundation, Santiago, Chile

@@

 Chile has experienced a persistent decrease in water supply, which impacts the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a mega-drought. There are few studies about the relationship between drought and ecosystem changes that can help with a better understanding of ecological drought. The aim of our study is to evaluate the interaction of drought, land cover change, and vegetation productivity over continental Chile. To assess drought, we used drought indices for atmospheric evaporative demand (AED), water supply, and soil moisture from short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales. We derived the drought indices using monthly ERA5-Land reanalysis data from 1981 to 2023. We used the Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive information on annual land cover and monthly vegetation productivity. Our results showed that except for the Austral part, Chile has a temporal decreasing trend in water supply, and across the whole country, there is an increase in AED. These trends become stronger over longer time scales. We found a negative trend in vegetation productivity in the north-central area, which is more prominent for shrubland and savanna as compared to croplands and forests. The anomaly in soil moisture over the past 12 months (SSI-12) is the most important variable explaining these changes, followed by anomalies in accumulated precipitation over one to two years (SPI-12 and SPI-24). The variable importance obtained by random forest models indicates that drought is explaining about 20–30% of the change in land cover surface across Chile for forest, grassland, shrubland, and savanna but has no relation to the changes in croplands. The increase in AED is the main variable associated with the change in land cover, followed by a reduction in precipitation and soil moisture. Our findings provide insightful information that could assist in developing adaptation measures for Chilean ecosystems to cope with climate change and drought. Also, this study could contribute to a better comprehension of ecological drought.

@@

# Introduction

Drought is often classified as 1) meteorological when precipitation in a specific period remains below the mean precipitation experienced in the same period during multiple years (more than 30 years usually), 2) hydrological when these anomalies last for long periods (months to years) and affect water systems, and 3) agricultural when the deficit negatively impacts plant health and leads to decreased productivity of crops or pastures [@Wilhite1985]. However, because drought is also influenced by human activities,  @Loon2016 and @AghaKouchak2021 expanded the drought definition for the Anthropocene, indicating that the feedback of human decisions and activities should also be considered (i.e.,  anthropogenic drought). Droughts lead to increased tree mortality [@Cheng2024] and induces alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017]. Even though many ecological studies have at times mistakenly considered "dry" conditions as "drought" [@Slette2019]. Ecological drought can be defined as _"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"_ [@Crausbay2017]. In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on future water security [@Bakker2012].

Global warming, as a result of human-induced greenhouse gas emissions, has increased the frequency and intensity of drought, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023]. The evidence supporting this claim has been strengthened since AR5 [@IPCC2013]. Recent studies, however, have produced contrasting findings, with some suggesting that drought has not exhibited a significant trend over the past forty years [@Vicente-Serrano2022; @Kogan2020]. @Vicente-Serrano2022 analyzed the trend in meteorological drought on a global scale, finding that only in a few regions an increase in the severity of drought was observed. Moreover, they attributed this increase solely to an increase in atmospheric evaporative demand (AED) due to higher temperatures, which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts. Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change, led by an increase in agricultural land"_. Similarly, @Kogan2020 analyzed the drought trend using remotely-sensed vegetation health indicators, finding that for the globe and main grain-producing countries, drought has not expanded or intensified during the past 38 years. Nonetheless, @IPCC2021 suggests that there is a medium to high degree of confidence that rising temperatures will increase the extent, frequency, and severity of agricultural and ecological droughts. Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. To better evaluate the impact of drought trends on ecosystems, assessments that correlate meteorological and soil moisture variables to their effects on vegetation are much needed.

From 1960 to 2019, land use change has impacted around one-third of the Earth’s surface, which is four times more than previously thought [@Winkler2021]. Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021;@Song2018] and regionally [@Chamling2020;@Homer2020;@Yang2021;@Schulz2010;@Echeverria2012]. Some seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012]. There is less research on drought and its relation to land cover change and vegetation productivity [@Chen2022;@Akinyemi2021;@Peng2017]. @Peng2017 utilized net primary productivity to examine the spatial and temporal variations in vegetation productivity at global level and assess to what extent drought influenced this variability by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change. According to their findings, drought is responsible for 37% of the decline and accounts for 55% of the variability in vegetation productivity. @Chen2022 instead found poor correlations (r<0.2) between the vegetation productivity trends against meteorological drought (SPEI of twelve months in December) and soil moisture at the global level.  These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) obtained for 12 month periods. SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow to know the contribution of each variable on its own. To better understand these contributions on land cover change and vegetation productivity the following questions may be asked: i) how do land cover and vegetation productivity respond to short- to long-term meteorological and soil moisture droughts? And ii) How is this response different between humid and arid climatic zones? Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

To address the previous questions over extensive regions, we can utilize gridded data on water availability, vegetation conditions, and the respective drought indices. For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012]. The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts. @Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporates the temperature effect by subtracting AED from precipitation. SPEI allows for analyzing the combined effect of precipitation and AED. Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2024]. Recently, there has been more interest in using AED to track droughts separately to better disengage the effects of precipitation from temperature-dependent effects [@Vicente-Serrano2020]. One of the reasons is that AED is linked to more flash droughts in limited water regions [@Noguera2022].  @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023]. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021]. @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus, many drought indices exist that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from Earth’s water balance (e.g., precipitation, AED, soil moisture). Climatic variability impacts vegetation development, with unfavorable conditions such as low precipitation and high temperatures usually promote a decrease in plant productivity. To monitor the response of vegetation for large areas, the common practice is to use satellite data. For example, the Normalized Difference Vegetation Index (NDVI) derived from frequent satellite observations of red and near infrared spectral reflectance, has been widely used as a proxy for biomass production [@Camps-Valls2021; @Paruelo2016;@Helman2014]. For Chile's cultivated land, @Zambrano2018 used the zcNDVI for assessing seasonal biomass production in response to drought. Comparing the various meteo-related and vegetation-based drought indices, we can further our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types [@Beck2023;@Luebert2022] make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega-drought [@Garreaud2017], which has impacted the Chilean ecosystem and consequently makes it highly vulnerable to climate change [@Barria2021;@AlvarezGarreton2021]. This mega-drought was defined by the annual time series of the Standardized Precipitation Index (SPI) at a time scale of twelve months at the end of each year (December) when having values below one standard deviation.  Some studies have addressed how this drought affects single ecosystems in terms of forest growth [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016].  The term "mega-drought" is used in Chile to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years. In Chile, the relationship between drought and the environment remains poorly understood. Hence, we aim to contribute to understanding how climatic and soil moisture droughts influence ecosystem dynamics in order to provide useful information that helps for a better understanding of ecological droughts and, at the same time, helps to make well-informed decisions on adaptation strategies.

@@

Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] ([@fig-studyArea]a), which determines its great ecosystem diversity [@Luebert2022] ([@fig-studyArea]c). The Andes Mountains are a main factor in climate variation [@Garreaud2009]. For an aggregated overview of the results of the study, we used the five Chilean macrozones: "Norte Grande," "Norte Chico," "Centro," "Sur," and "Asutral". "Norte Grande" (17°34’–25°42’S) and "Norte Chico" (25°42'-32°8'S) predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). In these two northern regions, the land is mostly bare, with a small surface of vegetation types such as shrubland and grassland. In the macrozones "Centro" (32°08'-36°12'S) and the northern half of "Sur," (36°12'-43°48'S) the main climate is Mediterranean, with warm to hot summers (Csa and Csb). Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%). An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral" (43°48'-56°00'S) Those zones have a large areal extent of forest and grassland. The southern part of the country has a tundra climate, and in "Austral," it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

To analyze land cover change, we used the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for each year from 2001 to 2022 and defines 17 classes (see Table S1). To maintain our focus on a large scale and follow the FAO classification [@FAO2022], we considered native and planted forests as "forests", which represent ecosystems dominated by larger trees. To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 Collection 6.1 from MODIS [@Didan2015]. MOD13A3 provides vegetation indices with 1km spatial resolution and monthly frequency. The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at https://lpdaa.usgs.gov/tools/data-pool/.

@@

For soil moisture, water supply, and water demand variables, we used ERA5L (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides the evolution of atmospheric and land variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3). \ref{tab-desEOD} shows a summary of the data and its main characteristics.

@@

$$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}
where

@@

To derive the drought indices of water supply and demand, soil moisture, and vegetation (i.e., the proxy of productivity), we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981–2023 and 2000–2023, respectively. The drought indices correspond to a historical anomaly of a variable (e.g., meteorological, vegetation, or soil moisture). To account for the anomaly, the common practice is to derive it following a statistical parametric method in which it is assumed that the statistical distribution of the data is known [@Heim2002]. A wrong decision in the statistical distribution is usually the highest source of uncertainty [@Laimighofer2022]. In the case of Chile, due to its high degree of climatic variability, it is difficult to choose a proper distribution without previous research that could be applicable throughout Chile. Here, we follow a non-parametric method for the calculation of the drought indices, in a similar manner as the framework proposed by @Farahmand2015.

@@

The $A^s_i$ corresponds to a moving window (convolution) that sums the variable for time scales $s$. Start from the last month (n) and sum the variable for s months, then follow month by month until the first month in which it could sum for s months (n-s+1). For example, using as a variable the precipitation, a period of twelve months (n), and a time scale of three months (s), it will be:

@@

Then, an inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$. Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(a_i)$ probabilities across a period of interest:

@@

The drought indices $SPI$, $SPEI$, $EDDI$, $SSI$, and $zcNDVI$ are obtained following the inverse normal approximation:

@@

For the proxy of vegetation productivity, we chose the time scale that best correlates with annual net primary productivity (NPP) across continental Chile. For this purpose, we calculated the zcNDVI for time scales of 1, 3, 6, and 12 months in December and compared it with the annual NPP. We used the NPP from the MOD17A3HGF [@Running2019] dataset (MODIS). We choose to use six months because the r-squared of zcNDVI with NPP highly increases from one to six months, but from six to 12 months has a lower improvement. We obtained an r-squared of 0.31 for forest and 0.72 for shrubland (refer to the supplementary material in Section S5). Then, we chose the proxy of vegetation productivity for six months, which we will name zcNDVI hereafter. It was calculated at a monthly frequency for 2000–2023.

@@

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen’s slope [@Sen1968]. Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers unlike regular regressions does, and it is a non-parametric method that is not influenced by the distribution of the data. We applied the Mann-Kendall test to see if the trend was significant and Sen's slope to estimate the magnitude of the trend. We did this for the indices SPI, EDDI, SPEI, and SSI using the six time scales with data from 1981 to 2023 (monthly frequency), resulting in 24 trends (per index and time scale). Then, we extracted the trend aggregated by each of the five macrozones: "Norte Grande" to "Asutral," and per land cover type: grassland, forest, cropland, shrubland, savanna, and barren land ([@fig-studyArea]d).

@@

To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has been previously used for studies of drought and land cover in Chile [@Fuentes2021;@Zambrano2018]. We regrouped the 17 classes into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023. We validate the land cover macroclasses regarding a highly detailed (30 m of spatial resolution) land cover map made for Chile by @Zhao2016 for 2013-2014. Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66. Section S2 in the Supplementary Material shows the procedure for validation.

We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023. After that, we calculated the trend's change in surface per land cover type and macroclass. We used Mann-Kendall for the significance of the trend [@Kendall1975] and Sen's slope to calculate the magnitude [@Sen1968].

To assess how water demand and supply, and soil moisture affect the variation in vegetation productivity across various land cover types, we avoid analyzing areas that have major land cover changes for 2021–2022.  To assess how zcNDVI varied irrespective of land cover change, we developed a persistence mask for land cover, which only retains pixels for which the macroclass remained the same for at least 80% of the years (2001–2022) ([@fig-studyArea]d).

@@

```{r fig-varImportance}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "cdcdcdc"
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/bars_relative_importance_RF.png")
```

>>>>

          country: 7500994, Chile.

@@

          city:  Enschede
          country: The Netherlands.

@@

          country: Chile.

@@

          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile
          country: Chile.

@@

        - id: ica3
          name: Instituto de Ciencias Agroalimentarias, Animales y Ambientales (ICA3), Universidad de O'Higgins
          city: San Fernando
          country: Chile.

@@

          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750
          country: Chile.

@@

          name: Data Observatory Foundation
          country: Chile.

@@

  Chile has experienced a persistent decrease in water supply, which impacts the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a mega-drought. There is yet insufficient understanding of ecological drought in Chile due to the limited studies on the relationship between drought and ecosystem changes. The aim of our study is to evaluate the interaction of drought, land cover change, and vegetation productivity over continental Chile. To assess drought, we used drought indices for atmospheric evaporative demand (AED), water supply, and soil moisture from short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales. We derived the drought indices using monthly ERA5-Land reanalysis data from 1981 to 2023. We used Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive information on annual land cover and monthly vegetation productivity. Our results showed that, except for the Austral part, Chile has a temporal decreasing trend in water supply, and across the whole country, there is an increase in AED. These trends become stronger over longer time scales. We found a negative trend in vegetation productivity in the north-central area, which is more prominent for shrubland and savanna as compared to croplands and forests. The anomaly in soil moisture over the past 12 months (SSI-12) is the most important variable explaining these changes, followed by anomalies in accumulated precipitation over one to two years (SPI-12 and SPI-24). The variable importance obtained by random forest models indicates that drought explains about 12–41% of the change in land cover surface across Chile for forest, grassland, shrubland, and savanna but has little relation to the changes in croplands. The increase in AED is the main variable associated with the change in land cover, followed by a reduction in precipitation and soil moisture. Our findings provide insightful information that could assist in developing adaptation measures for Chilean ecosystems to cope with climate change and drought.

@@

# Introduction
Drought can be classified as 1) meteorological, when precipitation in a specific period remains below the mean precipitation experienced in the same period during multiple years (more than 30 years usually); 2) hydrological, when these anomalies last for long periods (months to years) and affect water systems; and 3) agricultural, when the deficit negatively impacts plant health and leads to decreased productivity of crops or pastures [@Wilhite1985]. However, because drought is also influenced by human activities, @Loon2016 and @AghaKouchak2021 expanded the drought definition for the Anthropocene, indicating that the feedback of human decisions and activities should also be considered (i.e., anthropogenic drought). Droughts can lead to increased tree mortality [@Cheng2024] and induce alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017]. Even though many ecological studies have at times mistakenly considered "dry" conditions as "drought" [@Slette2019], ecological drought can be defined as _"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"_ [@Crausbay2017]. In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on future water security [@Bakker2012].

Global warming, as a result of human-induced greenhouse gas emissions, has increased the frequency and intensity of drought, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023]. The evidence supporting this claim has been strengthened since AR5 [@IPCC2013]. Recent studies, however, have produced contrasting findings, with some suggesting that drought has not exhibited a significant trend over the past forty years [@Vicente-Serrano2022; @Kogan2020]. @Vicente-Serrano2022 analyzed the trend in meteorological drought on a global scale, finding that only in a few regions an increase in the severity of drought was observed. Moreover, they attributed this increase solely to an increase in atmospheric evaporative demand (AED) due to higher temperatures, which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts. Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change, led by an increase in agricultural land"_. Similarly, @Kogan2020 analyzed the drought trend using remotely-sensed vegetation health indicators, finding that for the globe and main grain-producing countries, drought has not expanded or intensified during the past 38 years. Nonetheless, @IPCC2021 suggests that there is a medium to high degree of confidence that rising temperatures will increase the extent, frequency, and severity of agricultural and ecological droughts. Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. To better evaluate the impact of drought trends on ecosystems, assessments of the relationship between meteorological and soil moisture variables and their effects on vegetation are much needed.

From 1960 to 2019, land use change has impacted around one-third of the Earth’s surface, which is four times more than previously thought [@Winkler2021]. Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021;@Song2018] and regionally [@Chamling2020;@Homer2020;@Yang2021;@Schulz2010;@Echeverria2012]. Some seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012]. There is less research on drought and its relation to land cover change and vegetation productivity [@Chen2022;@Akinyemi2021;@Peng2017]. @Peng2017 utilized net primary productivity to examine the spatial and temporal variations in vegetation productivity at global level and assess to what extent drought influenced this variability by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change. According to their findings, drought is responsible for 37% of the decline and accounts for 55% of the variability in vegetation productivity. @Chen2022 instead found poor correlations (r<0.2) between the vegetation productivity trends against meteorological drought (SPEI of twelve months in December) and soil moisture at the global level. These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) obtained for 12 month periods. SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow to know the contribution of each variable on its own. To better understand these contributions on land cover change and vegetation productivity the following questions may be asked: i) how do land cover and vegetation productivity respond to short- to long-term meteorological and soil moisture droughts? And ii) how is this response different between humid and arid climatic zones? Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

To address the previous questions over extensive regions, we can utilize gridded data on water availability, vegetation conditions, and the respective drought indices. For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012]. The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts. @Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporates the temperature effect by subtracting AED from precipitation. SPEI allows for analyzing the combined effect of precipitation and AED. Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2024]. Recently, there has been more interest in using AED to track droughts separately to better disentangle precipitation from temperature-dependent effects [@Vicente-Serrano2020]. One of the reasons is that AED is more linked to flash droughts in water-limited regions [@Noguera2022]. @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023]. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021]. @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus, many drought indices exist that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from Earth's water balance (e.g., precipitation, AED, soil moisture). Climatic variability impacts vegetation development, with unfavorable conditions such as low precipitation and high temperatures usually promoting a decrease in plant productivity. To monitor the response of vegetation for large areas, the common practice is to use satellite data. For example, the Normalized Difference Vegetation Index (NDVI) derived from frequent satellite observations of red and near infrared spectral reflectance, has been widely used as a proxy for biomass production @Camps-Valls2021;@Paruelo2016;@Helman2014]. For Chile's cultivated land, @Zambrano2018 used the zcNDVI for assessing seasonal biomass production in response to drought. Comparing the various meteo-related and vegetation-based drought indices, we can further our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types [@Beck2023;@Luebert2022] make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega-drought [@Garreaud2017], which has impacted the Chilean ecosystem and consequently makes it highly vulnerable to climate change [@Barria2021;@AlvarezGarreton2021]. This mega-drought was defined by the annual time series of the Standardized Precipitation Index (SPI) at a time scale of twelve months at the end of each year (December) when having values below one standard deviation. Some studies have addressed how this drought affects single ecosystems in terms of forest growth [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018,@Zambrano2016]. The term "mega-drought" is used in Chile to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years. In Chile, the relationship between drought and the environment remains poorly understood. Hence, we aim to contribute to understanding how climatic and soil moisture droughts influence ecosystem dynamics in order to provide useful information that helps for a better understanding of ecological droughts and, at the same time, helps to make well-informed decisions on adaptation strategies.

@@

Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] [@fig-studyArea]a, which determines its great ecosystem diversity [@Luebert2022] ([@fig-studyArea]c). The Andes Mountains are a main factor in climate variation [@Garreaud2009]. For an aggregated overview of the results of the study, we used the five Chilean macrozones: "Norte Grande" (17°34'–25°42'S), "Norte Chico" (25°42'-32°8'S), "Centro" (32°08'-36°12'S), "Sur" (36°12'-43°48'S), and "Austral" (43°48'-56°00'S). "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico", the climate changes to an arid steppe with cold temperatures (Bsk). In these two northern regions, the land is mostly bare, with a small surface of vegetation types such as shrubland and grassland. In the macrozones "Centro" and the northern half of "Sur",  the main climate is Mediterranean, with warm to hot summers (Csa and Csb). Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%). An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral". Those zones have a large area of forest and grassland. The southern part of the country has a tundra climate, while "Austral" is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

To analyze land cover change, we used the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for each year from 2001 to 2022 and defines 17 classes (see Table S1). To maintain our focus on a large scale and follow the FAO classification [@FAO2022], we considered native and planted forests as "forests", which represent ecosystems dominated by larger trees. To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 Collection 6.1 from MODIS [@Didan2015]. MOD13A3 provides vegetation indices with 1km spatial resolution and monthly frequency. The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at [https://lpdaa.usgs.gov/tools/data-pool/](https://lpdaa.usgs.gov/tools/data-pool/).

@@

For soil moisture, water supply, and water demand variables, we used ERA5-Land (ERA5L) (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides the evolution of atmospheric and land variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3). Table \ref{tab-desEOD} shows a summary of the data and its main characteristics.

@@

$$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}

where:

@@

To derive the drought indices of water supply and demand, soil moisture, and vegetation (i.e., the proxy of productivity), we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981–2023 and 2000–2023, respectively. The drought indices correspond to a historical anomaly of a variable (e.g., meteorological, vegetation, or soil moisture). To account for the anomaly, the common practice is to derive it following a statistical parametric method in which it is assumed that the statistical distribution of the data is known [@Heim2002]. The use of an erroneous statistical distribution that does not fit the data is usually the highest source of uncertainty [@Laimighofer2022]. In the case of Chile, due to its high degree of climatic variability, it is difficult to choose a proper distribution without previous research that could be applicable throughout Chile. Here, we follow a non-parametric method for the calculation of the drought indices, in a similar manner as the framework proposed by @Farahmand2015.

@@

The $A^s_i$ corresponds to a moving window (convolution) that sums the variable for time scales $s$. This summation is done over $s$ months, starting from the most recent month ($n$) back in time until month $n-s+1$. For example, using as a variable the precipitation, a period of twelve months ($n$), and a time scale of three months ($s$), it will be:

@@

Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i^s$ to derive the $P(A_i^s)$ probabilities across a period of interest:

@@

An inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$. Thus, the drought indices $SPI$, $SPEI$, $EDDI$, $SSI$, and $zcNDVI$ and  obtained following the equation:

@@

For the proxy of vegetation productivity, we chose the time scale that best correlates with annual net primary productivity (NPP) across continental Chile. For this purpose, we calculated the zcNDVI for time scales of 1, 3, 6, and 12 months in December and compared it with the annual NPP. We used the NPP from the MOD17A3HGF [@Running2019] dataset (MODIS). We chose to use six months because the $R^2$ between zcNDVI and NPP reaches its highest value at six months. We obtained an $R^2$ of 0.31 for forest and 0.72 for shrubland (refer to the supplementary material in Section S5). Then, we chose the proxy of vegetation productivity for six months, which we will name zcNDVI hereafter. It was calculated at a monthly frequency for 2000–2023.

@@

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric Mann-Kendall test [@Kendall1975]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. Sen's slope has the advantage over normal regression that it is less affected by outliers, and as a non-parametric method it is not influenced by the distribution of the data. We applied the Mann-Kendall test to see if the trend was significant and Sen's slope to estimate the magnitude of the trend. We did this for the indices SPI, EDDI, SPEI, and SSI using the six time scales with data from 1981 to 2023 (monthly frequency), resulting in 24 trends (per index and time scale). Then, we extracted the trend aggregated by each of the five macrozones: "Norte Grande" to "Austral", and per land cover type: grassland, forest, cropland, shrubland, savanna, and barren land ([@fig-studyArea]d).

@@

To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 Collection 6.1 from MODIS. This product has been previously used for studies of drought and land cover in Chile [@Fuentes2021;@Zambrano2018]. We regrouped the 17 classes into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies (Table S1). Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023. We validate the land cover macroclasses regarding a highly detailed (30 m of spatial resolution) land cover map made for Chile by @Zhao2016 for 2013-2014. Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66. Section S2 in the Supplementary Material shows the procedure for validation.

We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2022. After that, we calculated the trend's change in surface per land cover type and macroclass. We used Mann-Kendall for the significance of the trend [@Kendall1975] and Sen's slope to calculate the magnitude [@Sen1968].

To assess how water demand and supply, and soil moisture affect the variation in vegetation productivity across various land cover types, we avoid analyzing areas that experienced major land cover changes in the 2021–2022 period. To assess how zcNDVI varied irrespective of land cover change, we developed a persistence mask for land cover, which only retains pixels for which the macroclass remained the same for at least 80% of the 22 years ([@fig-studyArea]d).

@@

```{=latex}
\begin{table}[!ht]
\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the $R^2$ reached by each random forest model. The white cells indicate that the landcover class is not significant in terms of surface area.}
\label{tab-landcoverTrendRF}
\includegraphics[width=0.5\textwidth]{../output/figs/table_r2_most_important_variables.png}
\end{table}
```
